---
title: "Accessing Data"
---

## Quick Start: Download Data

All processed datasets with cell type annotations are available on Zenodo.

### Automated Download (Recommended)

From the repository root, run:

```{r}
#| eval: false
source("data/downloadData.R")
downloadData()
```

This will download all four SCE/SpatialExperiment objects into your `data/covid/` and `data/merfish/` directories.

### Manual Download

Visit the [Zenodo repository](https://doi.org/10.5281/zenodo.18274942) and download:

-   `covid_data_sce.rds` — COVID-19 PBMC query (severe infection)
-   `normal_data_sce.rds` — Healthy PBMC reference
-   `dss9_data.rds` — MERFISH colitis query (day 9)
-   `healthy_data.rds` — Healthy colon MERFISH reference

Place files in:

```         
data/
├── covid/
│   ├── covid_data_sce.rds
│   └── normal_data_sce.rds
└── merfish/
├── dss9_data.rds
└── healthy_data.rds
```

## Data Contents

All objects are `SingleCellExperiment` or `SpatialExperiment` format with:

-   **Expression data** in `assays()` (counts and logcounts)
-   **Cell metadata** in `colData()` including:
    -   Ground truth cell type annotations (`tier2_merged`)
    -   SingleR predictions (`singler_annotations_merged`)
    -   Azimuth predictions (`azimuth_celltype_l1_merged`)
    -   CellTypist predictions (`celltypist_predicted_labels_merged`)
    -   scArches predictions (`scvi_prediction_merged`)
    -   Method-specific confidence/uncertainty scores
-   **Dimensionality reductions** in `reducedDims()` (PCA, UMAP)
-   **Spatial coordinates** (MERFISH only)

## Loading Data

Once downloaded:

```{r}
#| eval: false
library(SingleCellExperiment)

# COVID-19 data
covid_query <- readRDS("data/covid/covid_data_sce.rds")
covid_ref <- readRDS("data/covid/normal_data_sce.rds")

# MERFISH data
merfish_query <- readRDS("data/merfish/dss9_data.rds")
merfish_ref <- readRDS("data/merfish/healthy_data.rds")

# Inspect annotations
head(colData(covid_query))
```

## Dataset Details

### COVID-19 PBMC

-   **Query**: \~6,800 CD14+ monocytes from severe COVID-19 patients
-   **Reference**: \~23,200 CD14+ monocytes from healthy controls
-   **Total genes**: 18,472
-   **Source**: [Stephenson et al., Nature Medicine 2021](https://doi.org/10.1038/s41591-021-01329-2)
-   **Technology**: 10x Genomics scRNA-seq

### MERFISH Mouse Colitis

-   **Query**: 29,040 cells from DSS-treated colon (day 9, peak inflammation)
-   **Reference**: 27,140 cells from healthy colon (day 0)
-   **Total genes**: 943 (MERFISH panel)
-   **Source**: [MerfishData package](https://bioconductor.org/packages/MerfishData) (Geistlinger et al., 2024); original data from [Cadinu et al., Cell 2024](https://doi.org/10.1016/j.cell.2024.02.026)
-   **Technology**: MERFISH spatial transcriptomics

## Citation

If using these datasets, please cite:

**COVID-19 data:**

Stephenson, E., et al. (2021). Single-cell multi-omics analysis of the immune response in COVID-19. *Nature Medicine*, 27, 904–916.

**MERFISH data:**

Cadinu, P., et al. (2024). Charting the cellular biogeography in colitis reveals fibroblast trajectories and coordinated spatial remodeling. *Cell*, 187(8), 2010-2028.

Geistlinger, L., Moffitt, J., & Gentleman, R. (2024). MerfishData: Collection of public MERFISH datasets. R package version 1.8.0. [Bioconductor](https://bioconductor.org/packages/MerfishData)

**scDiagnostics package:**

[See main citation on home page](index.qmd)