---
title: "Accessing Data"
---

## Quick Start: Download Data

All processed datasets with cell type annotations are available on Zenodo.

### Automated Download (Recommended)

From the repository root, run:

```{r}
#| eval: false
source("data/downloadData.R")
downloadData()
```

This will download all four SCE/SpatialExperiment objects into your `data/covid/` and `data/merfish/` directories.

### Manual Download

Visit the [Zenodo repository](https://doi.org/10.5281/zenodo.18274942) and download:

- `covid_data_sce.rds` — COVID-19 PBMC query (severe infection)
- `normal_data_sce.rds` — Healthy PBMC reference
- `dss9_data.rds` — MERFISH colitis query (day 9)
- `healthy_data.rds` — Healthy colon MERFISH reference

Place files in:

data/
├── covid/
│   ├── covid_data_sce.rds
│   └── normal_data_sce.rds
└── merfish/
├── dss9_data.rds
└── healthy_data.rds

## Data Contents

All objects are `SingleCellExperiment` or `SpatialExperiment` format with:

- **Expression data** in `assays()` (logcounts)
- **Cell metadata** in `colData()` including:
  - Ground truth cell type annotations (`tier2_merged`)
  - SingleR predictions (`singler_annotations_merged`)
  - Azimuth predictions (`azimuth_celltype_l1_merged`)
  - CellTypist predictions (`celltypist_predicted_labels_merged`)
  - scArches predictions (`scvi_prediction_merged`)
  - Method-specific confidence/uncertainty scores
- **Dimensionality reductions** in `reducedDims()` (UMAP, PCA)

## Loading Data

Once downloaded:

```{r}
#| eval: false
library(SingleCellExperiment)

# COVID-19 data
covid_query <- readRDS("data/covid/covid_data_sce.rds")
covid_ref <- readRDS("data/covid/normal_data_sce.rds")

# MERFISH data
merfish_query <- readRDS("data/merfish/dss9_data.rds")
merfish_ref <- readRDS("data/merfish/healthy_data.rds")

# Inspect annotations
head(colData(covid_query))
```

## Dataset Details

### COVID-19 PBMC

- **Query**: 6,765 CD14+ monocytes from severe COVID-19 patients
- **Reference**: 6,804 CD14+ monocytes from healthy controls
- **Source**: [Ren et al., Cell 2021](https://doi.org/10.1016/j.cell.2021.02.018)
- **Technology**: 10x Genomics scRNA-seq

### MERFISH Mouse Colitis

- **Query**: 44,324 cells from DSS-treated colon (day 9, peak inflammation)
- **Reference**: 26,482 cells from healthy colon (day 0)
- **Source**: [Cadinu et al., Cell 2024](https://doi.org/10.1016/j.cell.2024.02.026)
- **Technology**: MERFISH spatial transcriptomics
- **Genes**: 943 genes measured

## Citation

If using these datasets, please cite:

**COVID-19 data:**
Ren, X., et al. (2021). COVID-19 immune features revealed by a large-scale single-cell transcriptome atlas. *Cell*, 184(7), 1895-1913.

**MERFISH data:**
Cadinu, P., et al. (2024). Charting the cellular biogeography in colitis reveals fibroblast trajectories and coordinated spatial remodeling. *Cell*, 187(8), 2010-2028.

**scDiagnostics package:**
[See main citation on home page](index.qmd)
```

