---
title: "COVID-19 PBMC Analysis"
---

## Overview

This page demonstrates the application of `scDiagnostics` to the COVID-19 PBMC data, comparing single-cell measurements from severe COVID-19 patients to a healthy reference. We use a combination of project mapping in reduced dimensional space, anomaly detection, and differential expression analysis to illustrate how scDiagnostics can facilitate the detection and characterization of a disease-associate cell state that would be missed by standard reference-based annotation tools.

## Dataset Context

- **Healthy Reference:** 23,201 PBMC cells from 23 healthy patients
- **COVID-19 Query:** 48,148 PBMC cells from 86 severe COVID-19 patients
- **Focus:** Interferon (IFN) response genes and transcriptomic shifts during severe infection

## Reproducibility Across Annotation Methods

**All results below are reproducible with any annotation method.**

Results shown use Azimuth L1 annotations (`azimuth_celltype_l1_merged`), but identical analyses can be generated by simply changing the `query_cell_type_col` parameter to:

- `"singler_annotations_merged"` — SingleR correlation-based annotations
- `"celltypist_predicted_labels_merged"` — CellTypist machine learning predictions
- `"scvi_prediction_merged"` — scArches deep learning transfer

We consistently observe the same patterns of anomaly detection, ECM dysregulation, and elevated IFN gene expression across all four methods, confirming that results reflect genuine disease-driven transcriptomic remodeling rather than artifacts from individual annotationtools.

## Analysis Components

### 1. UMAP Visualization (Figure A)

**Script:** `R/covid/UMAP Code (scVI).R`

Three complementary UMAP visualizations of the joint reference-query space:

**A) Disease Status UMAP (Left)**

- Healthy reference cells (blue) cluster tightly in the lower left region
- COVID-19 query cells (pink) form a separate, larger cluster to the right
- Clear spatial segregation indicates substantial transcriptomic remodeling during severe infection

**B) Cell Type UMAP (Center)**

- Cell type annotations are preserved across both datasets
- Reference landscape: B cells, Plasma cells, HSPCs, and diverse T/NK populations
- Immune composition remains similar despite disease state

**C) IFN Signature UMAP (Right)**

- Dramatic difference in interferon-stimulated gene expression
- Reference cells: predominantly low IFN scores (dark purple)
- Query cells: predominantly high IFN scores (yellow/orange)
- This uniform, elevated IFN activation marks COVID-19 cells as a distinct transcriptomic state

**Note:** Alternative dimensionality reduction methods (Fast MNN and Harmony) are also available in `UMAP Code (Fast MNN).R` and `UMAP Code (Harmony).R`.

### 2. PCA with IFN Signature (Figure B)

**Script:** `R/covid/PCA IFN Plot.R`

Pairwise scatterplot matrix of first 3 PCs with:

- **Diagonal:** Density distributions for reference (blue) vs. query (purple)
- **Lower panels:** PCA scatter plots colored by IFN signature score
  - Reference cells (hollow circles) cluster tightly in negative PC space with low IFN scores
  - Query cells (solid circles) scatter across positive PC space with elevated IFN scores
- **Upper panels:** Blank for clarity

**Left panel (Reference):** Homogeneous reference population with minimal variation; IFN activation is baseline

**Right panel (Query):** Heterogeneous distribution of COVID-19 cells; IFN activation drives separation from reference in PC space

### 3. Anomaly Detection (Figure B, Right)

**Script:** `R/covid/PCA Anomaly Plot.R`

Identifies cells with transcriptomic profiles substantially different from the reference:

- **Anomalous cells** (red): ~50% of COVID-19 cells with transcriptional patterns not well-represented in healthy reference
- **Non-anomalous cells** (green/black): ~50% with relatively normal transcriptional structure
- Reference cells (shown left) have minimal anomalies

**Key finding:** A substantial and systematic fraction of COVID-19 cells are flagged as anomalous, indicating genuine disease-driven transcriptomic remodeling beyond simple continuous variation.

::: {.callout-note collapse="true"}
## Function Details: detectAnomaly()

```{r}
#| eval: false
#| code-fold: false

anomaly_output <- detectAnomaly(
    query_data = covid_data,
    reference_data = normal_data,
    query_cell_type_col = "azimuth_celltype_l1_merged",
    ref_cell_type_col = "author_cell_type_merged",
    cell_types = c("CD14 mono"),
    pc_subset = 1:5,
    n_tree = 500,
    anomaly_threshold = 0.5,
    assay_name = "logcounts",
    max_cells_query = NULL,
    max_cells_ref = NULL
)

# Extract results for CD14 monocytes
query_anomaly_status <- anomaly_output[["CD14 mono"]][["query_anomaly"]]
reference_anomaly_status <- anomaly_output[["CD14 mono"]][["reference_anomaly"]]
```

**Parameters:**
- `query_data`, `reference_data` — SCE objects to compare
- `query_cell_type_col`, `ref_cell_type_col` — Column names containing cell type labels
- `cell_types` — Specific cell type(s) to analyze
- `pc_subset` — Principal components to use (1:5 = first 5 PCs)
- `n_tree` — Number of trees for isolation forest (500 = balance speed/accuracy)
- `anomaly_threshold` — Score threshold for anomaly classification (0.5 = 50th percentile)

**Output:** Logical vectors indicating which cells are anomalous in each dataset

**Flexibility note:** You can change `query_cell_type_col` to use different annotation systems:
- `"azimuth_celltype_l1_merged"` (shown here)
- `"singler_annotations_merged"`
- `"celltypist_predicted_labels_merged"`
- `"scvi_prediction_merged"`

We observe consistent anomaly detection patterns across all four annotation methods, confirming that results reflect genuine biological divergence rather than method-specific artifacts.
:::

### 4. Gene Expression Shifts (Figure C)

**Script:** `R/covid/Heatmap and Barplot Code.R`

Analyzes 25 interferon-stimulated genes to quantify transcriptomic shifts:

**C) Heatmap (Left)**

- **Rows:** 25 interferon-stimulated genes (IFN signature)
- **Columns:** Three comparisons (gray brackets at top)
  - Query vs Reference (all)
  - Query non-anomalous vs Reference
  - Query anomalous vs Reference
- **Color:** Blue (low) to Red (high expression)
- Reference cells show uniformly low expression (blue)
- Query cells show dramatically elevated expression (red)
- Anomalous query cells show the most extreme upregulation

**C) Barplot (Right)**

- Individual log2-fold changes for each IFN gene
- Gray bars: All query vs reference (median ~0.8 log2-fold)
- Red bars: Anomalous vs reference (median ~1.5 log2-fold)
- Light red bars: Non-anomalous vs reference (median ~0.5 log2-fold)
- Nearly all 25 genes show statistically significant upregulation

**Key observations:**

- **Uniform IFN activation:** All 25 genes consistently upregulated; no selective subset
- **Anomaly stratification:** Anomalous cells show 2-3 fold greater expression increase than non-anomalous cells
- **Strongest signals:** LY6E, ISG15, and MX1 show the most dramatic fold changes (>1.5 log2-fold in anomalous)
- **Statistical significance:** Paired t-tests show highly significant differences (p < 0.001) between anomalous and non-anomalous populations; Cohen's d indicates large effect sizes

::: {.callout-note collapse="true"}
## Function Details: calculateGeneShifts()

```{r}
#| eval: false
#| code-fold: false

gene_shifts <- calculateGeneShifts(
    query_data = covid_data[yoshida_ifn_signature, ],
    reference_data = normal_data[yoshida_ifn_signature, ],
    query_cell_type_col = "azimuth_celltype_l1_merged",
    ref_cell_type_col = "author_cell_type_merged",
    cell_types = c("CD14 mono"),
    pc_subset = 1:5,
    n_top_loadings = 30,
    assay_name = "logcounts",
    p_value_threshold = 0.05,
    adjust_method = "fdr",
    detect_anomalies = TRUE,
    anomaly_comparison = TRUE,
    anomaly_threshold = 0.5,
    max_cells_query = 5000,
    max_cells_ref = 5000
)

# Plot results
plot(
    gene_shifts,
    cell_type = "CD14 mono",
    pc_subset = 1:5,
    plot_type = "heatmap",
    plot_by = "p_adjusted",
    n_genes = 25,
    significance_threshold = 0.05,
    show_anomalies = TRUE,
    pseudo_bulk = TRUE,
    max_cells_query = 5000,
    max_cells_ref = 5000
)
```

**Parameters:**
- `query_data`, `reference_data` — SCE objects (subset to genes of interest)
- `query_cell_type_col`, `ref_cell_type_col` — Cell type labels
- `cell_types` — Specific cell type to analyze
- `pc_subset` — Principal components determining gene importance
- `n_top_loadings` — Top genes by PC loadings to consider
- `detect_anomalies` — Include anomaly detection (TRUE/FALSE)
- `anomaly_comparison` — Separate results by anomaly status (TRUE/FALSE)
- `max_cells_*` — Subsample for computational efficiency

**Output:** Gene shift statistics with p-values, fold changes, and anomaly-stratified comparisons

**Flexibility note:** All four annotation systems show similar gene shift patterns. Changing `query_cell_type_col` to any of the merged columns produces consistent results, validating that observed transcriptomic shifts are robust biological signals independent of annotation method.
:::

## Interpretation

**COVID-19 cells exhibit:**

1. **Spatial separation** — Distinct clustering from healthy reference in embedding space
2. **Uniform IFN activation** — All 25 interferon genes co-regulated; reflects coordinated antiviral response
3. **High anomaly rates** — High % of cells flagged as transcriptomically divergent from reference
4. **Stratified dysregulation** — Anomalous cells show extreme IFN upregulation (1.5-2.0 log2-fold) vs. non-anomalous cells (~0.5 log2-fold)

**Implications for annotation:**

- **Distance-based methods may struggle** with cells showing this degree of transcriptomic shift
- **Anomaly detection provides actionable signal** — Identifies the most dysregulated cells independent of annotation tool
- **Built-in confidence scores may be insufficient** — Uniform IFN activation suggests systematic, coordinated remodeling that confidence scores may underestimate
- **Biological signal is robust** — Consistent across multiple diagnostic approaches (PCA, anomaly detection, gene shifts) and reproducible with all four annotation methods

## Figures Generated

- `scvi_umap_disease.png` — UMAP by disease status
- `scvi_umap_celltype.png` — UMAP by cell type
- `scvi_umap_ifn.png` — UMAP by IFN signature
- `pca_ifn.png` — PCA pairs plot with IFN scores and anomaly detection
- `anomaly_plot.png` — PCA anomaly detection results
- `covid_ifn_heatmap.png` — Gene shift heatmap
- `gene_shifts_barplot.png` — Gene shift barplot


