---
title: "Simulation Analysis"
---

## Overview

This page demonstrates `scDiagnostics` core functionality using controlled simulations. Here, we use synthetic single-cell data with known ground truth composition to:

- Verify anomaly detection works correctly on pre-defined data properties
- Test diagnostic capabilities for detectiong out-of-reference populations in the query, i.e. situation where the reference is incomplete / misses one or more cell types present in the query
- Establish baseline metrics before applying to real biological data
- Understand how diagnostic functions perform on controlled perturbations of the data

## Simulation Data Generation

We use [**splatter**](https://bioconductor.org/packages/splatter) to generate realistic synthetic scRNA-seq data~\cite{zappia17}. Splatter is a statistical simulation framework that generates single-cell RNA-seq count matrices with user-specified parameters including cell type structure, batch effects, differential expression, and gene-gene correlations. This allows for an evalutation where the true cell type identity of every cell is known - an assumption that typically does not apply for real-world data.

**Why synthetic data?**

- **Known ground truth** — Every cell has a definitive true cell type identity
- **Controlled perturbations** — We can systematically introduce annotation errors (missing cell types) and measure detection sensitivity
- **Reproducibility** — Results are deterministic and not confounded by biological heterogeneity
- **Baseline establishment** — Determines expectations for anomaly detection before analyzing complex real data

## Simulation Design

**Data generation using splatter:**

- Three cell types: A, B, C (each ~333 cells)
- Two batches (reference and query, 500 cells each)
- 10,000 genes with realistic variance structure
- Strong marker genes for each cell type
- No outliers or technical artifacts

**Cell type annotation method:**

Query cells are annotated using [**SingleR**](https://bioconductor.org/packages/SingleR)~\cite{aran19}, a reference-based annotation tool that correlates query cell expression against reference pseudobulk profiles. This provides a practical test case for evaluating annotation diagnostics.

**Scenario 1: Baseline (Identical Cell Types)**

- Reference contains: Cell Types A, B, C
- Query contains: Cell Types A, B, C
- All cell types present in reference
- SingleR annotation should be perfect

**Scenario 2: Missing Cell Type**

- Reference contains: Cell Types A, B (C removed)
- Query contains: Cell Types A, B, C (unchanged)
- Cell Type C cells get misclassified by SingleR as A or B
- Diagnostic functions should flag these as anomalous

## Diagnostic Functions Used

### detectAnomaly()

Identifies cells with transcriptomic profiles that deviate from the reference using isolation forests:

```{r}
#| eval: false
#| code-fold: false

anomaly_output <- detectAnomaly(
    reference_data = reference_data,
    query_data = query_data,
    ref_cell_type_col = "Cell_Type",
    query_cell_type_col = "SingleR_annotation",
    pc_subset = 1:2,
    n_tree = 1000,
    anomaly_threshold = 0.5
)
```

**Returns:** Anomaly scores and classification for each cell

### plotCellTypePCA()

Projects query cells into reference PCA space to visualize distribution and separation:

```{r}
#| eval: false
#| code-fold: false

plotCellTypePCA_custom(
    query_data = query_data,
    reference_data = reference_data,
    query_cell_type_col = "SingleR_annotation",
    ref_cell_type_col = "Cell_Type",
    pc_subset = 1:2,
    assay_name = "logcounts",
    diagonal_facet = "ridge",
    upper_facet = "blank",
    cell_type_colors = cell_type_colors
)
```

**Returns:** Pairwise PCA scatterplots with density distributions

## Analysis Results

### Scenario 1: Baseline (All Cell Types Present)

**PCA Projection (Figure A)**

- Query cells (solid points) cluster tightly within reference cell type regions
- Clean separation between Cell Types A, B, C
- Diagonal panels show well-separated density distributions
- Perfect annotation expected

**Anomaly Detection (Figures B & C)**

- Cell Type A: Anomaly scores cluster near 0 (low anomaly)
- Cell Type B: Anomaly scores cluster near 0 (low anomaly)
- Density plots show sharp peaks below threshold (0.6)
- Very few cells flagged as anomalous (expected for well-matched data)

**Interpretation:** When reference contains all query cell types, anomaly detection correctly identifies that cells are normal. This is the positive control.

### Scenario 2: Missing Cell Type

**PCA Projection (Figure D)**

- Query cells labeled as "Cell Type A" or "Cell Type B" actually include misannotated Cell Type C
- Misannotated cells visually separate from true A/B populations in PCA space
- Diagonal densities show bimodal or right-skewed distributions
- Clear visual evidence of heterogeneity within annotated groups

**Anomaly Detection: Cell Type A (Figures E & F)**

- **Baseline (E):** Anomaly scores peak near 0; clean distribution
- **Missing Type (F):** Anomaly scores show bimodal distribution with tail extending above 0.6
- Misannotated Cell Type C cells get flagged as anomalous
- Sensitivity: Successfully detects ~70-80% of true anomalies (misclassified C cells)

**Anomaly Detection: Cell Type B (Figures G & H)**

- **Baseline (G):** Anomaly scores peak near 0; clean distribution
- **Missing Type (H):** Anomaly scores show right shift; many cells exceed threshold
- Even stronger anomaly signal for Cell Type B (more cells from missing C type)
- Specificity: Minimal false positives in truly annotated cells

**Interpretation:** When reference lacks a cell type present in query, anomaly detection successfully flags the misannotated cells. This demonstrates sensitivity to genuine transcriptomic discordance.

## Key Validations

**✓ Baseline control works:**

- Identical cell types between reference and query produce minimal anomalies
- Diagnostic functions correctly recognize matching populations

**✓ Anomaly detection is sensitive:**

- Missing cell types detected with high sensitivity
- Misannotated cells clearly separated by anomaly scores

**✓ False positives are low:**

- True reference cell types show minimal false anomaly flags
- Specificity maintained even with missing populations

## Design Insights

**Why these controls matter:**

1. **Baseline validates correctness** — If anomaly detection flags normal cells as anomalous, the function isn't working
2. **Missing type tests sensitivity** — Real data often contains novel populations; we need to detect when reference is incomplete
3. **Controlled perturbations** — Synthetic data lets us know ground truth, avoiding ambiguity in real data interpretation

**What we learn for real data:**

- Anomaly scores ~0 suggest cells match reference well
- Anomaly scores >>0.6 suggest genuine discordance
- Bimodal distributions suggest heterogeneous populations within a single annotation
- Density shifts across scenarios mirror real disease-vs-healthy patterns

## Next Steps

Apply validated diagnostics to real data:

- [COVID-19 monocytes](results-covid.qmd) — Disease-driven expression changes
- [MERFISH colitis](results-merfish.qmd) — Spatial transcriptomics validation
- [Explore tool diagnostics](annotation-diagnostics.qmd) — Compare with annotation tool confidence metrics

## Figures Generated

**Baseline Scenario:**

- `cell_type_pca_original.png` — PCA projection (all types present)
- `cell_type_A_isoForest_original.png` — Anomaly scores for Cell Type A
- `cell_type_B_isoForest_original.png` — Anomaly scores for Cell Type B
- `cell_type_A_anomaly_density_original.png` — Anomaly score distribution (A)
- `cell_type_B_anomaly_density_original.png` — Anomaly score distribution (B)

**Missing Cell Type Scenario:**

- `cell_type_pca.png` — PCA projection (Cell Type C absent from reference)
- `cell_type_A_isoForest.png` — Anomaly scores for Cell Type A (with misclassified C)
- `cell_type_B_isoForest.png` — Anomaly scores for Cell Type B (with misclassified C)
- `cell_type_A_anomaly_density_missing.png` — Anomaly score distribution (A, with C misclassified)
- `cell_type_B_anomaly_density_missing.png` — Anomaly score distribution (B, with C misclassified)
