---
title: "Setup & Installation"
---

## Overview

This page provides instructions for setting up the computational environment needed to run the analysis tutorials and preprocessing pipelines. The workflow requires both R and Python packages.

## Quick Start

If you're only downloading pre-processed data and running tutorials, you mainly need **R packages**. If you're preprocessing data from scratch, you'll need the full **R + Python** setup.

## Part 1: R Packages

All R dependencies can be installed with a single setup script.

### Installation Script

Run this R script once to install all required packages:

```{r}
#| eval: false
source("R/covid/R Package Installation Pipeline.R")
```

Or for MERFISH-specific analysis:

```{r}
#| eval: false
source("R/merfish/R Package Installation Pipeline.R")
```

**Note:** Both scripts install largely the same packages. Running either one will set you up for both COVID-19 and MERFISH analyses.

### What Gets Installed

**From CRAN:**
- Data manipulation: `dplyr`, `tidyr`, `tibble`
- Matrix operations: `Matrix`
- Utilities: `reticulate`, `remotes`, `here`, `ggplot2`, `patchwork`, `viridis`

**From Bioconductor:**

- Core: `SingleCellExperiment`, `SpatialExperiment`
- Analysis: `scran`, `scater`, `SingleR`, `biomaRt`
- I/O: `zellkonverter`, `HDF5Array`, `MerfishData`
- Parallel: `BiocParallel`

**From GitHub:**

- `Seurat` (v5.3.1.0 - specific version for Azimuth compatibility)
- `Azimuth` (reference-based annotation)
- `scDiagnostics` (anomaly detection)

### Checking Installation

After running the script, verify everything installed:

```{r}
#| eval: false
# Test loading key packages
library(SingleCellExperiment)
library(scran)
library(Azimuth)
library(scDiagnostics)

cat("✓ All R packages loaded successfully!\n")
```

---

## Part 2: Python Environment (Optional)

Only needed if you're running CellTypist or scVI/scArches annotation from scratch.

### Option A: scVI/scArches via Conda

This is the **recommended approach** for the complete pipeline.

**Step 1:** Create the conda environment

```bash
conda env create -f environment-scvi.yml
```

**Step 2:** Activate it when running Python scripts

```bash
conda activate scvi-env
```

The `environment-scvi.yml` file is in the repository root and includes:

- Python 3.9
- `scvi-tools` and `scarches` (annotation tools)
- `scanpy` (data handling)
- JAX (GPU support, optional)

### Option B: CellTypist Setup

If using CellTypist separately, run the Python environment setup script:

```{r}
#| eval: false
source("R/auxiliary/environmentSetupCellTypist.R")
environmentSetupCellTypist()
```

This function:

- Detects your Python installation
- Creates a `celltypist_env` conda environment
- Installs required packages: `scanpy`, `pandas`, `numpy`, `celltypist`
- Tests imports and provides manual installation instructions if needed

### Manual Python Installation

If you prefer manual setup:

```bash
# For scVI/scArches
conda create -n scvi-env python=3.9
conda activate scvi-env
pip install scvi-tools scarches scanpy leidenalg scikit-learn

# For CellTypist
conda create -n celltypist_env python=3.9
conda activate celltypist_env
conda install -c conda-forge scanpy pandas numpy
pip install celltypist
```

---

## System Requirements

- **R:** Version 4.2 or later
- **Python:** Version 3.9 (recommended for compatibility)
- **Memory:** 16+ GB RAM recommended
- **Disk:** ~50 GB for raw data + processed outputs

## Troubleshooting

**R package installation fails:**

- Ensure you have C++ compilers installed (`Rtools` on Windows, `build-essential` on Linux, Xcode on Mac)
- Update R to the latest stable version
- Try installing packages individually with `BiocManager::install("package_name")`

**Python environment issues:**

- Ensure conda is installed: `conda --version`
- Clear conda cache: `conda clean --all`
- Reinstall environment: `conda env remove -n scvi-env && conda env create -f environment-scvi.yml`

**reticulate can't find Python:**

In R, explicitly set the Python path:

```{r}
#| eval: false
library(reticulate)
use_condaenv("scvi-env")  # or "celltypist_env"
```

---

## Next Steps

Once installation is complete:

1. **Just want to explore results?** → Go to [Getting Data](data-access.qmd)
2. **Want to preprocess data?** → Follow [COVID-19 Processing](processing-covid.qmd) or [MERFISH Processing](processing-merfish.qmd)
3. **Ready for analysis?** → Start with [scDiagnostics Overview](overview-scdiagnostics.qmd)
