---
title: "Setup & Installation"
---

## Overview

This page provides instructions for setting up the computational environment needed to run the preprocessing pipelines and analysis tutorials. The workflow requires both R and Python packages.

## Quick Start

If you're primarily interested in running the tutorials on pre-processed data (fast), you only need the **R dependencies**. If you also want to preprocess the data from scratch (slow), you will need to complete the full **R + Python** setup.

## Part 1: R Packages

All R dependencies can be installed with a single setup script. The package requirements are largely identical for both COVID-19 and MERFISH analyses.

### Installation Script

Run this R script once to install all required packages:

```{r}
#| eval: false
# For COVID-19 scRNA-seq analysis
source("R/covid/R Package Installation Pipeline.R")
```

Or equivalently for MERFISH colitis analysis:

```{r}
#| eval: false
# For MERFISH spatial transcriptomics analysis
source("R/merfish/R Package Installation Pipeline.R")
```

**Note:** Both scripts install the same core packages with identical versions. Running either one will set you up for both analyses.

### What Gets Installed

**From CRAN (15 packages):**

- Data manipulation: `dplyr`, `tidyr`, `tibble`  
- Matrix operations: `Matrix`  
- Visualization: `ggplot2`, `cowplot`, `patchwork`, `GGally`, `ggridges`, `pheatmap`, `circlize`, `viridis`
- Utilities: `reticulate`, `remotes`, `here`

**From Bioconductor (14 packages):**

- Infrastructure: `SingleCellExperiment`, `SpatialExperiment`, `HDF5Array`, `DelayedArray`, `BiocParallel`, `BiocSingular`
- Analysis: `scran`, `scater`, `SingleR`
- Visualization: `ComplexHeatmap`
- Utilities: `biomaRt`, `zellkonverter`, `MerfishData`  
- Diagnostics: `scDiagnostics` (devel v1.5.1)

**From GitHub (2 packages):**

- `Seurat` (v5.3.1.0 - specific version for Azimuth compatibility)
- `Azimuth` (reference-based annotation)

### Checking Installation

After running the script, verify everything is installed correctly:

```{r}
#| eval: false
# Test loading key packages
library(SingleCellExperiment)
library(scran)
library(Azimuth)
library(scDiagnostics)

cat("✓ All R packages installed and loaded successfully!\n")
```

---

## Part 2: Python Environment (Optional)

Only needed if you run CellTypist or scVI/scArches annotation from scratch.

### Option A: scVI/scArches via Conda (GPU-Accelerated)

This is the **recommended approach** for the complete preprocessing pipeline.

**Important:** scVI/scArches is computationally intensive and is designed to run on GPU hardware. We performed all analyses on GPU nodes (NVIDIA L40S GPUs on HMS O2 cluster). CPU-only execution will be significantly slower.

**Step 1:** Create the conda environment

```bash
conda env create -f environment-scvi.yml
```

**Step 2:** Activate it when running Python scripts

```bash
conda activate scvi-env
```

The `environment-scvi.yml` file is in the repository root and includes:

- Python 3.9
- `scvi-tools` and `scarches` (GPU-enabled annotation tools)
- `scanpy` (data analysis)
- JAX (with GPU support)
- CUDA Toolkit integration

**GPU Access:** Ensure your compute environment has GPU access. Example SLURM submission on O2:

```bash
sbatch --gres=gpu:l40s:1 --mem=40G your_script.py
```

### Option B: CellTypist Setup

For using CellTypist separately (CPU-compatible), run the Python environment setup script:

```{r}
#| eval: false
source("R/auxiliary/environmentSetupCellTypist.R")
environmentSetupCellTypist()
```

The `environmentSetupCellTypist()` function:

- detects your Python installation
- creates a `celltypist_env` conda environment
- installs required packages: `scanpy`, `pandas`, `numpy`, `celltypist`
- tests imports and provides manual installation instructions if needed

### Manual Python Installation

If you prefer manual setup:

```bash
# For scVI/scArches (GPU-enabled)
conda create -n scvi-env python=3.9
conda activate scvi-env
pip install scvi-tools scarches scanpy leidenalg scikit-learn jax jaxlib
```

```bash
# For CellTypist (CPU-compatible)
conda create -n celltypist_env python=3.9
conda activate celltypist_env
conda install -c conda-forge scanpy pandas numpy
pip install celltypist
```

---

## System Requirements

- **R:** Version 4.2 or later
- **Python:** Version 3.9 (recommended for compatibility)
- **Memory:** 16+ GB RAM for R, 40+ GB RAM for GPU-enabled scVI/scArches
- **GPU:** NVIDIA GPU with CUDA 11.8+ (required for scVI/scArches)
- **Disk:** ~50 GB for raw data + processed outputs

## Troubleshooting

**R package installation failure:**

- Ensure you have C++ compilers installed (`Rtools` on Windows, `build-essential` on Linux, Xcode on Mac)
- Update R to the latest release version
- Try installing packages individually with `BiocManager::install("package_name")`

**Python environment issues:**

- Ensure conda is installed: `conda --version`
- Clear conda cache: `conda clean --all`
- Reinstall environment: `conda env remove -n scvi-env && conda env create -f environment-scvi.yml`
- Verify GPU access: `nvidia-smi` (if running on HPC cluster, check with `module list`)

**reticulate can't find Python:**

In R, explicitly set the Python path:

```{r}
#| eval: false
library(reticulate)
use_condaenv("scvi-env")  # or "celltypist_env"
```
